The Strucplot Framework:\\ Visualizing Multi-way Contingency Tables with \pkg{vcd}

### 2006-09-20 trunk/vcd/paper/Strucplot/JSS.Rnw
- spacings-default synchronized - pairs_diagonal_mosaic added - labeling can now create alternating labels
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@571 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

\noindent Current implementations of mosaic displays can be found,
e.g., for \proglang{SAS} \citep{vcd:SAS:2005}, \proglang{ViSta} \citep{vcd:young:1996}, 
\proglang{MANET} \citep{vcd:Unwin+Hawkins+Hofmann:1996},
\proglang{Mondrian} \citep{vcd:Theus:2003},
\proglang{R} \citep{vcd:R:2005}, and \proglang{S-PLUS} \citep{vcd:SPLUS:2005}. 
Table \ref{tab:compare} gives an overview of the available functionality in these systems.

\begin{table}[h]
  \centering
  \begin{tabular}{|l|c|c|c|c|c|}
    \hline
                             & \proglang{SAS} & \proglang{S-PLUS} & \proglang{R} & \pkg{ViSta} & \pkg{MANET}/\pkg{Mondrian}\\\hline
    Basic functionality      &  $\times$  &   $\times$    & $\times$ &   $\times$   & $\times$\\
    Shape                    &     &        & $\times$ &       & $\times$\\
    Residual-based shadings  &  $\times$  &        & $\times$ &   $\times$   & ($\times$)\\
    Conditional Views        &  $\times$  &        & $\times$ &       & $\times$\\
    Interaction              &     &        &   &   $\times$   & $\times$\\
    Extensible Design        &     &        & $\times$ &       &  \\\hline
  \end{tabular}
  \caption{Comparison of current software environments.}
  \label{tab:compare}
\end{table}

@@

\noindent The small bullets indicate zero entries in the corresponding
cell. Note that in contrast to, e.g.,  \codefun{mosaicplot} in base
\proglang{R}, the default split direction and level ordering in all strucplot
displays correspond to the textual representation.
It is also possible to visualize the expected values instead of the
observed values (see Figure \ref{fig:expected}):

@@

By a suitable combination of splitting, spacing, and
labeling settings, the functions provided by the strucplot framework
can be customized in a quite flexible way. For example, 
\codefun{doubledecker} is simply a wrapper for 
\codefun{mosaic}, setting the right defaults. 
Figure \ref{fig:doubledecker} shows a doubledecker plot of
the \data{Titanic} data, explaining the probability of survival (`survived') by age, given
sex, given class. It is created by:

@@

@@

Spacing of strucplot displays is customizable in a similar way than shading. The
\code{spacing} argument of the \codefun{strucplot} function takes a
list of \class{unit} vectors, one for each dimension, specifying the
space between the tiles corresponding to the levels. 
Consider again the introductory example of the \data{Arthritis} data 
(Figure \ref{fig:arthritis}). Since we are interested in the effect of
the medicament in the placebo and treatment groups, a mosaic plot is
certainly appropriate to visualize the three levels of `Improved' in
the two `Treatment' strata. Another conceptual approach is to use
spine plots with highlighting \citep{vcd:hummel:1996}. A spine plot is
a variation of a bar plot where the heights of the bars are held
constant, whereas the widths are used to represent the number of cases
in each category. This is equivalent to a mosaic plot for a one-way table.
If a second (indicator) variable is highlighted in a spine plot, we
obtain a display equivalent to a simple mosaic display for a two-way
table, except that no space between the levels of the highlighted
variable is used.
In the \data{Arthritis} example, we will highlight patients with `Marked' 
improvement in both groups. To obtain such a display within 
the strucplot framework\footnote{The \pkg{vcd} package also provides 
a specialized \codefun{spine} function for spine plots and spinograms.}, it
suffices to set the space between the `Improved' tiles to 0 (see
Figure \ref{fig:artspine}):

@@

>>>>

\noindent Current implementations of mosaic displays can be found,
e.g., for \proglang{SAS} \citep{vcd:SAS:2005}, \proglang{ViSta} \citep{vcd:young:1996}, 
\proglang{MANET} \citep{vcd:Unwin+Hawkins+Hofmann:1996},
\proglang{Mondrian} \citep{vcd:Theus:2003},
\proglang{R} \citep{vcd:R:2006}, and \proglang{S-PLUS}
\citep{vcd:SPLUS:2005}. For \proglang{R}, currently three
implementations do exist in the packages \pkg{graphics} (in base \proglang{R}), 
\pkg{vcd} \citep{vcd:Meyer+Zeileis+Hornik:2006},
and \pkg{iPlots} \citep{vcd:urbanek+wichtrey:2006}, respectively.
Table \ref{tab:compare} gives an overview of the available functionality in these systems.
Most environments are available on Windows, Apple Macintosh, and Linux/Unix
variants, except \proglang{MANET} which is only available for the
Macinthosh platforms. \proglang{Mondrian} is a JAVA application, as
well as major parts of the \pkg{iPlots} package.

\begin{table}[h]
  \centering
  \begin{tabular}{|l|c|c|c|c|c|c|}
    \hline
                             & \proglang{SAS} & \proglang{S-PLUS} &
                             \proglang{R} & \pkg{ViSta} & \pkg{MANET}
                             & \pkg{Mondrian}\\\hline
    Basic functionality      &  $\times$  &   $\times$    & $\times$ &   $\times$   & $\times$& $\times$\\
    Shape                    &     &        & $\times$ & $\times$  &  $\times$&\\
    Residual-based shadings  &  $\times$  &        & $\times$ &   highlighting   &($\times$)& ($\times$)\\
    Conditional Views        &  $\times$  &        & $\times$ & $\times$ &  $\times$&\\
    Interaction              &     &        & $\times$  &   $\times$   & $\times$& $\times$\\
    Linking                  &     &        & $\times$  &       & $\times$& $\times$\\
    Extensible Design        &     &        & $\times$ &       &  &\\
    Language                 & SAS & S & S,JAVA & XLisp & C++ & JAVA\\
\hline
  \end{tabular}
  \caption{Comparison of current software environments.}
  \label{tab:compare}
\end{table}

%%DM%% Fixme: separate columns for iplot, base, and vcd?

@@

\noindent For zero entries, tiles have zero area and are, additionally,
marked by small bullets (see, e.g, Figure \ref{fig:doubledecker}). By
default, these cells are not splitted further. The bullets help
distinguishing very small cells from zero entries, and are
particularly useful when color shadings come into play 
(see the example using the \data{Bundesliga} data in Section \ref{sec:overview}). 
Note that in contrast to, e.g.,  \codefun{mosaicplot} in base
\proglang{R}, the default split direction and level ordering in all strucplot
displays correspond to the textual representation.
It is also possible to visualize the expected values instead of the
observed values (see Figure \ref{fig:expected}):

@@

By a suitable combination of splitting, spacing, and
labeling settings, the functions provided by the strucplot framework
can be customized in a quite flexible way. For example, 
\codefun{doubledecker.default} is simply a wrapper for 
\codefun{strucplot}, setting the right defaults:

<<doubledecker>>=
doubledecker.default <- 
  function(x,
           depvar = length(dim(x)), 
           margins = c(1, 4, length(dim(x)) + 1, 1),
           col = rainbow_hcl(tail(dim(x), 1), end = 260),
           labeling = labeling_doubledecker,
           spacing = spacing_highlighting,
           main = NULL, 
           keep_aspect_ratio = FALSE,
           ...) {
  d <- dim(x)
  l <- length(d)
  if (is.character(depvar))
    depvar <- match(depvar, names(dimnames(x)))
  condvars <- (1:l)[-depvar]
  x <- aperm(x, c(condvars, depvar))
  strucplot(x,
         condvars = l - 1,
         spacing = spacing,
         split = c(rep.int(TRUE, l - 1), FALSE),
         gp = gpar(fill = col),
         shade = TRUE,
         labeling = labeling,
         main = main,
         margins = margins,
         legend = NULL,
         keep_aspect_ratio = keep_aspect_ratio,
         ...
         )
}
@

\noindent Most default settings such as colors, spacing, and labeling are specified via the parameters of
\codefun{doubledecker.default} and passed through to
\codefun{strucplot}. The additional code just handles the dependent
variable information, and in particular permutes the table to have the
variable as the last dimension as required for the doubledecker plot.
Figure \ref{fig:doubledecker} shows a doubledecker plot of
the \data{Titanic} data, explaining the probability of survival (`survived') by age, given
sex, given class. It is created by:

@@

\noindent (Note that the bullets drawn for zero cells are shaded, too).

@@

Spacing of strucplot displays is customizable in a similar way than shading. The
\code{spacing} argument of the \codefun{strucplot} function takes a
list of \class{unit} vectors, one for each dimension, specifying the
space between the tiles corresponding to the levels. 
Consider again the introductory example of the \data{Arthritis} data 
(Figure \ref{fig:arthritis}). Since we are interested in the effect of
the medicament in the placebo and treatment groups, a mosaic plot is
certainly appropriate to visualize the three levels of `Improved' in
the two `Treatment' strata. Another conceptual approach is to use
spine plots with highlighting \citep{vcd:hummel:1996}. A spine plot is
a variation of a bar plot where the heights of the bars are held
constant, whereas the widths are used to represent the number of cases
in each category. This is equivalent to a mosaic plot for a one-way table.
If a second (indicator) variable is highlighted in a spine plot, we
obtain a display equivalent to a simple mosaic display for a two-way
table, except that no space between the levels of the highlighted
variable is used.
In the \data{Arthritis} example, we will highlight patients with `Marked' 
improvement in both groups. To obtain such a display within 
the strucplot framework, it suffices to set the space between the `Improved' tiles to 0 (see
Figure \ref{fig:artspine}):

@@

\noindent Note that the default and formula methods for 
\codefun{mosaic} provide a
convenience interface for highlighting. A similar plot (with slightly
different shading) than the previous one can be obtained using:

<<artspine>>=
mosaic(Improved ~ Treatment, data = Arthritis, split_vertical = TRUE)
@
